---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1



## Loading and preprocessing the data
Load and preprocess the data as follows.

1. Load any required libraries. 
2. Assume the data exists in the current working directory as "activity.zip". Unzip the data file if needed.
3. Load the resulting .csv file into R. Note that the missing data is coded as NA, so set na.strings = "NA", and set strinsAsFactors = FALSE to read in dates as strings instead of factors. 
4. Convert the date column to dates using R package lubridate.
```{r}
   library(lubridate)
   library(dplyr)
   library(ggplot2)

    # unzip the file if needed
    if (!file.exists("./foo.txt")) {
        unzip("./activity.zip", exdir="./")
    }
    # list directory contents -- extracted file is "activity.csv"
    dir()
    # load the .csv into R and convert date coulmn to dates
    act <- read.csv("./activity.csv", stringsAsFactors = FALSE)
    act$date <- ymd(act$date)
    # call str() to check the format of the data frame
    str(act)
```



## What is mean total number of steps taken per day?
For the initial analysis, ignore missing values. 

Use dplyr group_by() to group the data by day and summarize() to get the daily step totals.
```{r}
    act_by_date <- group_by(act, date)
    daily_step_totals <- summarize(act_by_date, "Total.Steps" = sum(steps, na.rm = TRUE))
    
    # check the min and max to help determine number of bins
    min(daily_step_totals$Total.Steps)
    max(daily_step_totals$Total.Steps)
    
```

Calculate the mean and median daily steps.
```{r}
    ds_median <- median(daily_step_totals$Total.Steps, na.rm = TRUE)
    ds_mean <- round(mean(daily_step_totals$Total.Steps, na.rm = TRUE),2)
```

Use ggplot2 to plot a histogram of daily step totals. 
```{r}
    d <- ggplot(daily_step_totals, aes(Total.Steps))
    d + geom_histogram(binwidth = 500) +
      xlab("Daily Step Totals") +
      ylab("Number of Days") +
      ggtitle("Daily Step Totals, 10/1/2012 - 11/30/2012") +
      geom_vline(xintercept=ds_median, color = "red", size = 1) + 
      geom_vline(xintercept=ds_mean, color = "blue", size = 1) +
      annotate("text",label=paste("Median = ",ds_median), x=17500, y=9, size=4, color = "red") +
      annotate("text",label=paste("Mean = ",ds_mean), x=17500, y=7, size=4, color = "blue")
```

The median daily step total is `r ds_median`.

The mean daily step total is `r ds_mean`.

## What is the average daily activity pattern?
Use dplyr group_by() to group the data by interval and summarize() to get the average steps per interval.
```{r}
    act_by_int <- group_by(act, interval)
    int_step_avg <- summarize(act_by_int, "Average.Steps" = mean(steps, na.rm = TRUE))
```

Compute the max value and the interval which contains the max.
```{r}
    as_max <- round(max(int_step_avg$Average.Steps, na.rm = TRUE),2)
    as_max_idx <- which.max(int_step_avg$Average.Steps)
    as_max_int <- int_step_avg$interval[as_max_idx]
```

Use ggplot to make a time series plot of average steps vs. interval.
```{r}
    i <- ggplot(int_step_avg, aes(y=Average.Steps,x=interval))
    i + geom_line() +
      xlab("Interval") +
      ylab("Interval Step Average") +
      ggtitle("Interval Step Averages, 10/1/2012 - 11/30/2012") +
      geom_vline(xintercept=as_max_int, color = "red", size = 1) + 
      #geom_vline(xintercept=ds_mean, color = "blue", size = 1) +
      annotate("text",label=paste("Max = ", as_max, "\n@ Interval = ",as_max_int), x=2000, y=150, size=4, color = "red")
    
```

Interval `r as_max_int` contains the maximum interval step average of `r as_max`.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
